---
title: "Lab5"
format: html
author: "Gabriel Castellanos"
---

## Introduction

This project calls for loading 3 different files (XML, JSON, and, HTML) that each describe 3 of my favorite books and their attributes. Each file was created with Notepad ++ and loaded into a Github Repository. In the end, we should have 3 seperate data frames.

## JSON

We need jsonlite package

```{r}
install.packages("jsonlite")
library(jsonlite)

url <- "https://raw.githubusercontent.com/gc521/DATA-607-Data-Acquisition-Mangement/Lab-5/Books.json"

book <-fromJSON(url)

as.data.frame(book)


```

## XML

This can be done with the aid of few packages.

```{r}
install.packages("xml2")
library(xml2)
install.packages("XML")
library("XML")
install.packages('methods')
library(methods)
result <- read_xml('https://raw.githubusercontent.com/gc521/DATA-607-Data-Acquisition-Mangement/Lab-5/Books.xml')

books.xml <- xmlParse(result)

books.xml <- xmlToDataFrame(books.xml)


```
## HTML

```{r}
install.packages("rvest")
library(rvest)


books.html <- read_html('https://raw.githubusercontent.com/gc521/DATA-607-Data-Acquisition-Mangement/Lab-5/Books.html', trim = TRUE)
books.html <- html_table(books.html)
books.html <- as.data.frame(books.html)


```

## Conclusion

All 3 file types posed their own unique challenges, but if you know which packages to use, the task becomes very manageable. 